---
title: "Socrates_coACR_tutorial"
author: "Alexandre P. Marand"
date: "11/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Quick preprocessing

Clean, normalize and cluster the `Socrates` object that is automatically loaded into the namespace with the package. In this example, we will use `tfidf` for a bit of a speed-up over other available options.

```{r process}

# load Socrates
library(Socrates)

# filter, normalize, and reduce dimensions
obj <- cleanData(obj, min.c=1000, min.t=0.001, max.t=0.005)
obj <- tfidf(obj)
obj <- reduceDims(obj, n.pcs=50, cor.max=0.7)
obj <- projectUMAP(obj)

# cluster
obj <- callClusters(obj, res=0.6)

```



## Identify co-accessible ACRs across all cells

To find co-accessible ACRs, `Socrates` needs a path to the reference genome. A chromosome lengths file for hg19 is provided in `Socrates`. 

```{r}

# get genome annotation
genome <- system.file("extdata", "hg19.txt", package="Socrates")
```


`Socrates` relies on the R package `cicero`, developed by the Trapenll and Shendure labs to identify co-accessible ACRs. Please cite their work [here](https://www.cell.com/molecular-cell/fulltext/S1097-2765(18)30547-1#%20) if using `Socrates` to identify co-accessible regions. For more information and documentation for `cicero`, visit the software website site [here](https://cole-trapnell-lab.github.io/cicero-release/docs/#references).

```{r coACRs-default, eval=FALSE}

# find co-accessible ACRs in the entire data set (not subset by cluster)
obj <- coAccess(obj, genome=genome)

```



## Identify co-accessible ACRs per cluster

Identifying co-accessible ACRs within each cluster independently is as simple as specifying an additional parameter in the function `coAccess`. Co-accessible ACRs are identified by each group in the `groupID` column of the meta data table `obj$Clusters` slot. The default `groupID` is "LouvainClusters", which is the default column name for the function `callClusters`. See `?callClusters` and `?coAccess` for more details.

```{r coACRs-group, eval=FALSE}

# find co-accessible ACRs by cluster
obj <- coAccess(obj, genome=genome, byGroup=T)

```



## Empirical FDR to control false positives

Removing false positive links based solely on the degree of co-accessibility is a challenge because different clusters/cell types are afflicted by varying degrees of technical variation, such as read depth. A simple approach to estimate the false discovery rate is to shuffle the ACR x barcode matrix, keeping the number of accessible ACRs per cell and the number of accessible cells per ACR identical to the original matrix. Co-accessible ACR identification is repeated and heuristic thresholds removing 95% (or other FDRs of the users choice) of co-accessible ACRs from the shuffled matrix (for positive and negatively associated ACRs) are identified. These thresholds are then applied to the original, non-shuffled matrix, resembling FDR controlled co-accessible ACRs. Users can mix and match the various approaches described here. An example is illustrated below.

```{r coACRs-eFDR}

# find co-accessible ACRs with eFDR
obj <- coAccess(obj, genome=genome, eFDR=TRUE, fdr_thresh=0.05)

# co-accessible ACRs are stored in the 'obj$coACR' slot. 
head(obj$coACRs)

```
