---
title: "Socrates_coACR_tutorial"
author: "Alexandre P. Marand"
date: "11/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Quick preprocessing

Clean, normalize and cluster the `Socrates` object that is automatically loaded into the namespace with the package. In this example, we will use `tfidf` for a bit of a speed-up over other available options.

```{r process}

# filter, normalize, and reduce dimensions
obj <- cleanData(obj, min.c=1000, min.t=0.001, max.t=0.005)
obj <- tfidf(obj)
obj <- reduceDims(obj, n.pcs=50, cor.max=0.7)
obj <- projectUMAP(obj)

# cluster
obj <- callClusters(obj, res=0.6)

```

## Identify co-accessible ACRs

To find co-accessible ACRs, `Socrates` needs a few annotation files, including a path to the reference genome and gene coordinates. Below is a quick example of how to identify co-accessible ACRs using all barcodes.

```{r coACRs-1}

# find co-accessible ACRs in the entire data set (not subset by cluster)
obj <- coAccess(obj, genome=)

```